### 2025-07-17 paper/quarto_aaai/paper.tex
shorted one sentence
Draft adjustments in discussions/conclusions

\iffalse
%Example, Multiple Authors, ->> remove \iffalse,\fi and place them surrounding AAAI title to use it
% \title{My Publication Title --- Multiple Authors}
\author {
    % Authors
    First Author Name\textsuperscript{\rm 1},
    Second Author Name\textsuperscript{\rm 2},
    Third Author Name\textsuperscript{\rm 1}
}
\affiliations {
    % Affiliations
    \textsuperscript{\rm 1}Affiliation 1\\
    \textsuperscript{\rm 2}Affiliation 2\\
    firstAuthor@affiliation1.com, secondAuthor@affilation2.com, thirdAuthor@affiliation1.com
}
\fi
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amsthm}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\usepackage{placeins}
\usepackage{siunitx}
\sisetup{uncertainty-mode = separate}

@@

Specifically, \textbf{we propose Counterfactual Training (CT)}: a novel
training regime explicitly geared towards improving the explanatory
capacity of models that, in high-level terms, we define as the extent to
which valid counterfactual explanations derived for a model can be
deemed plausible with respect to the underlying data and global
actionability constraints (we refine this notion in Def.
\ref{def-explainability}). To the best of our knowledge, Counterfactual
Training represents the first attempt to achieve more explainable models
by employing counterfactual explanations already in the training phase.

The remainder of this manuscript is structured as follows.
Section~\ref{sec-lit} presents related work, focusing on the link
between AEs and CEs. Then follow our two principal contributions. In
Section~\ref{sec-method}, we introduce our methodological framework and
show theoretically that it can be employed to enforce global
actionability constraints. In Section~\ref{sec-experiments}, through
extensive experiments, we empirically demonstrate that CT substantially
improves explainability and positively contributes to the robustness of
trained models without sacrificing predictive performance. Finally, in
Section~\ref{sec-discussion}, we discuss open challenges and conclude
that CT is a promising approach towards making opaque models more
trustworthy.

@@

In the context of counterfactual explanations, plausibility has broadly
been defined as the degree to which generated CEs comply with the
underlying data-generating process
\citep{altmeyer2024faithful, guidotti2022counterfactual, poyiadzi2020face}.
Plausibility is a necessary but insufficient condition for using CEs to
provide algorithmic recourse (AR) to individuals (negatively) affected
by opaque models. An AR recommendations must also be actionable, i.e.,
possible to attain by the recipient. A plausible CE for a rejected
20-year-old loan applicant, for example, might reveal that their
application would have been accepted, if only they had been 20 years
older. Ignoring all other features, this would comply with the
definition of plausibility if 40-year-old individuals were in fact more
credit-worthy on average than young adults. But to the applicant, this
CE does not provide actionable recourse, as age is not a (directly)
mutable feature. Counterfactual training aims to improve model
explainability by aligning models with counterfactuals that meet both
desiderata: plausibility and actionability. Formally, we define
explainability as follows:

@@

In all four cases, it is possible to generate valid counterfactuals
(stars) for unsuccessful applicants (blue). They cross the decision
boundary (green) into the target class, but their quality differs. In
panel (a), they are not plausible: they do not comply with the
distribution of the factuals in \(y^+\) to the point where they form a
clearly discernible cluster. In panel (b), they are highly plausible,
meeting the first objective of Def. \ref{def-explainability}. This
difference in outcomes is quantified for the non-linear MLP in the first
two columns of Table~\ref{tbl-main} as the \(\%\)-reduction in
implausibility: it is substantial and statistically significant for
\emph{LS} across both metrics, the distance-based \(\text{IP}\) (29\%)
and divergence-based \(\text{IP}^*\) (55\%).

@@

We find that CT generally leads to substantial and statistically
significant improvements in plausibility: average reductions in
\(\text{IP}\) range from around 7\% for \emph{MNIST} to almost 60\% for
\emph{Circ}; for the real-world tabular datasets they are around 12\%
for both \emph{CH} and \emph{Cred} and almost 25\% for \emph{GMSC}; for
\emph{Adult} and \emph{OL} we find no significant impact of CT on
\(\text{IP}\). Reductions in \(\text{IP}^*\) are even more substantial
and generally statistically significant, although the average degree of
uncertainty is higher than for \(\text{IP}\): average reductions range
from around 20\% (\emph{Moons}) to almost 90\% (\emph{Circ}). The only
negative findings for \emph{OL} and \emph{MNIST} are statistically
insignificant and, for MNIST, do not align with qualitative findings,
which are much more plausible for CT (Figure~\ref{fig-mnist}).

@@

We also find that CT can reduce sensitivity to immutable, protected
features and thus lead to less costly counterfactual outcomes as shown
in Figure~\ref{fig-poc}. In column three of Table~\ref{tbl-main}, we
impose mutability constraints on selected features and compute the
reduction in average costs of CEs associated with CT compared to the
baseline: for synthetic datasets, we always protect the first feature;
for all real-world tabular datasets we could identify and protect an
\emph{age} variable; for \emph{MNIST}, we protect the five upper and
lower pixel rows of the full image. Reductions in costs are
overwhelmingly positive and significant of up to nearly 60\% for
\emph{GMSC}. While the estimated cost reductions for \emph{Adult} and
\emph{MNIST} are not significant, Figure~\ref{fig-mnist} (columns 6-10)
demonstrates that CT does have the expected effect: sensitivity to
protected features as per integrated gradients is drastically reduced;
details of this experiment are reported in the supplementary appendix.
In the case of \emph{Cred}, average costs increase, likely because any
potential benefits from protecting the \emph{age} are outweighed by the
increase in costs required for greater plausibility.

@@

\caption{\label{tbl-main}Key performance metrics and bootstrapped
standard errors (omitted if essentially zero) for all datasets.
\textbf{Plausibility} (columns 1-2): percentage reduction in
implausibility for \(\text{IP}\) and \(\text{IP}^*\), respectively;
\textbf{Actionability} (column 3): percentage reduction in costs with
protected features. \textbf{Accuracy} (columns 4-7): test accuracies and
robust accuracies (\(\text{Acc}^*\)) for CT and the baseline (BL).
Counterfactual outcomes in columns 1-3 are aggregated across bootstrap
samples and varying degrees of the energy penalty
\(\lambda_{\text{egy}}\) used for \emph{ECCo} at test time. Asterisks
(\(^*\)) in columns 1-3 indicate that difference is statistically
significant at 1\%-level. Standard errors for accuracy are bootstrapped
from the test set.}

@@

\section{Conclusions}\label{sec-discussion}

@@

\emph{Interventions on features have implications for fairness.} We
provide a tool that allows practitioners to modify the sensitivity of a
model with respect to certain features. Model owners can use our
solution to support the fair and equitable treatment of decision
subjects, but they could also misuse it by enforcing explanations based
on features that are more difficult to modify by some (group of)
individuals. When used irresponsibly, CT could result in an unfairly
assigned burden of recourse \citep{sharma2020certifai}, threatening the
equality of opportunity \citep{bell2024fairness}. Additionally, CT
requires mutability constraints for the features considered by the
model. Even if all immutable features are protected, there may exist
proxies that are mutable, and hence should not be protected, but
preserve sufficient information about the principals to hinder these
protections. Deciding on actionability is still a major open challenge
in the AR literature \citep{venkatasubramanian2020philosophical}
impacting the capacity of CT to fulfill its intended goal.

\emph{CT increases the training times.} Like adversarial training, CT is
more resource-intensive than conventional regimes. Higher numbers of CEs
improve the quality of learned representations but they also increase
the number of computations. As our codebase is not performance
optimized, grids of 270 settings for the largest datasets in our
experiments took up to four hours using 34 2GB CPUs (see supplementary
appendix). Other than optimization, three factors mitigate this effect:
(1) CT yields itself to parallel execution; (2) it amortizes the cost of
CEs for the training samples; and (3) it can be used to fine-tune
conventionally-trained models.

We also highlight three important directions for future research.
Firstly, it is an interesting challenge to extend CT beyond
classification settings. Our formulation relies on the distinction
between non-target class(es) \(y^{-}\) and target class(es) \(y^{+}\) to
generate counterfactuals through Equation~\ref{eq-obj}. While \(y^{-}\)
and \(y^{+}\) can be arbitrarily defined, CT requires the output space
\(\mathcal{Y}\) to be discrete. Thus, it does not apply to ML tasks
where the change in outcome cannot be readily discretized. Focus on
classification is a common choice in research on CEs and AR. Other
settings have attracted some interest, e.g., regression
\citep{spooner2021counterfactual}, but there is little consensus how to
robustly extend the notion of CEs.\\
Secondly, CT is susceptible to training instabilities. This problem has
been recognized for JEMs \citep{grathwohl2020your} and even though we
depart from the SGLD sampling, we still encounter variability in
outcomes. CT is exposed to two potential sources of instabilities: (1)
the energy-based contrastive divergence term, \(\text{div}(\cdot)\), in
Equation~\ref{eq-div}, and (2) the underlying explainers. We find
several promising ways to mitigate this problem: regularizing energy
(\(\lambda_{\text{reg}}\)), generating sufficiently many counterfactuals
during each epoch, and including only mature counterfactuals in
\(\text{div}(\cdot)\).\\
Finally, we believe that it is possible to considerably improve
hyperparameter selection procedures, and thus performance. We have
relied exclusively on grid searches, but future research could benefit
from more sophisticated approaches.

To conclude, state-of-the-art machine learning models are prone to
learning complex representations that cannot be interpreted by humans.
Existing explainability solutions cannot guarantee that explanations
agree with the model's learned representation of data. As a step towards
addressing this challenge, we introduced counterfactual training, a
novel training regime that incentivizes highly-explainable models. Our
approach leads to explanations that are both plausible (compliant with
the underlying data-generating process) and actionable (compliant with
user-specified mutability constraints), and thus meaningful to their
recipients. Through extensive experiments, we demonstrated that CT
satisfies its objective while promoting robustness and preserving the
predictive performance of models. It can also be used to fine-tune
conventionally-trained models and achieve similar gains. Lastly, our
work highlights the value of simultaneously improving models and their
explanations.

>>>>

\iffalse
%Example, Multiple Authors, ->> remove \iffalse,\fi and place them surrounding AAAI title to use it
% \title{My Publication Title --- Multiple Authors}
\author {
    % Authors
    First Author Name\textsuperscript{\rm 1},
    Second Author Name\textsuperscript{\rm 2},
    Third Author Name\textsuperscript{\rm 1}
}
\affiliations {
    % Affiliations
    \textsuperscript{\rm 1}Affiliation 1\\
    \textsuperscript{\rm 2}Affiliation 2\\
    firstAuthor@affiliation1.com, secondAuthor@affilation2.com, thirdAuthor@affiliation1.com
}
\fi
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amsthm}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\usepackage{placeins}
\usepackage{siunitx}
\sisetup{uncertainty-mode = separate}
\DeclareMathSizes{10}{9}{7}{6.5}

@@

Specifically, \textbf{we propose Counterfactual Training (CT)}: a novel
training regime explicitly geared towards improving the explanatory
capacity of models that, in high-level terms, we define as the extent to
which valid explanations derived for a model can be deemed plausible
with respect to the underlying data and global actionability constraints
(we refine this notion in Def. \ref{def-explainability}). For
simplicity, we refer to models with high explanatory capacity as
\emph{explainable}. To the best of our knowledge, Counterfactual
Training represents the first attempt to achieve more explainable models
by employing counterfactual explanations already in the training phase.

The remainder of this manuscript is structured as follows.
Section~\ref{sec-lit} presents related work, focusing on the link
between AEs and CEs. Then follow our two principal contributions. In
Section~\ref{sec-method}, we introduce our methodological framework and
show theoretically that it can be employed to enforce global
actionability constraints. In Section~\ref{sec-experiments}, through
extensive experiments, we empirically demonstrate that CT substantially
improves explainability and positively contributes to the robustness of
trained models without sacrificing predictive performance. In
Section~\ref{sec-discussion}, we discuss open challenges and, in
\textbf{?@sec-conclusion}, conclude that CT is a promising approach
towards making opaque models more trustworthy.

@@

In the context of counterfactual explanations, plausibility has broadly
been defined as the degree to which generated CEs comply with the
underlying data-generating process
\citep{altmeyer2024faithful, guidotti2022counterfactual, poyiadzi2020face}.
Plausibility is a necessary but insufficient condition for using CEs to
provide algorithmic recourse (AR) to individuals (negatively) affected
by opaque models. An AR recommendations must also be actionable, i.e.,
possible to attain by the recipient. A plausible CE for a rejected
20-year-old loan applicant, for example, might reveal that their
application would have been accepted, if only they had been 20 years
older. Ignoring all other features, this would comply with the
definition of plausibility if 40-year-old individuals were in fact more
credit-worthy on average than young adults. But of course this CE does
not qualify for providing actionable recourse to the applicant since
\emph{age} is not a (directly) mutable feature. Counterfactual training
aims to improve model explainability by aligning models with
counterfactuals that meet both desiderata: plausibility and
actionability. Formally, we define explainability as follows:

@@

In all four cases, it is possible to generate valid counterfactuals
(stars) for unsuccessful applicants (blue). They cross the decision
boundary (green) into the target class, but their quality differs. In
panel (a), they are not plausible: they do not comply with the
distribution of the factuals in \(y^+\) to the point where they form a
clearly discernible cluster. In panel (b), they are highly plausible,
meeting the first objective of Def. \ref{def-explainability}. This
difference in outcomes is quantified for the non-linear MLP in the first
two columns of Table~\ref{tbl-main} as the \(\%\)-reduction in
implausibility: it is substantial and statistically significant for
\emph{LS} across both metrics, the distance-based \(IP\) (29\%) and
divergence-based \(IP^{*}\) (55\%).

@@

We find that CT generally leads to substantial and statistically
significant improvements in plausibility: average reductions in \(IP\)
range from around 7\% for \emph{MNIST} to almost 60\% for \emph{Circ};
for the real-world tabular datasets they are around 12\% for both
\emph{CH} and \emph{Cred} and almost 25\% for \emph{GMSC}; for
\emph{Adult} and \emph{OL} we find no significant impact of CT on
\(IP\). Reductions in \(IP^{*}\) are even more substantial and generally
statistically significant, although the average degree of uncertainty is
higher than for \(IP\): average reductions range from around 20\%
(\emph{Moons}) to almost 90\% (\emph{Circ}). The only negative findings
for \emph{OL} and \emph{MNIST} are statistically insignificant and, for
MNIST, do not align with qualitative findings, which are much more
plausible for CT (Figure~\ref{fig-mnist}).

@@

We also find that CT can reduce sensitivity to immutable, protected
features and thus lead to less costly counterfactual outcomes as shown
in Figure~\ref{fig-poc}. In column three of Table~\ref{tbl-main}, we
impose mutability constraints on selected features and compute the
reduction in average costs of CEs associated with CT compared to the
baseline: for synthetic datasets, we always protect the first feature;
for all real-world tabular datasets we could identify and protect an
\emph{age} variable; for \emph{MNIST}, we protect the five top and
bottom rows of digits. Reductions in costs are overwhelmingly positive
and significant of up to nearly 60\% for \emph{GMSC}. While the
estimated cost reductions for \emph{Adult} and \emph{MNIST} are not
significant, Figure~\ref{fig-mnist} (columns 6-10) demonstrates that CT
does have the expected effect: sensitivity to protected features as per
integrated gradients is drastically reduced; details of this experiment
are reported in the supplementary appendix. In the case of \emph{Cred},
average costs increase, likely because any potential benefits from
protecting the \emph{age} are outweighed by the increase in costs
required for greater plausibility.

@@

\caption{\label{tbl-main}Key performance metrics and bootstrapped
\emph{two} standard errors for all datasets. \textbf{Plausibility}
(columns 1-2): percentage reduction in implausibility for \(IP\) and
\(IP^*\), respectively; \textbf{Actionability} (column 3): percentage
reduction in costs with protected features. \textbf{Accuracy} (columns
4-7): test accuracies and robust accuracies (\(\text{Acc}^*\)) for CT
and the baseline (BL). Counterfactual outcomes in columns 1-3 are
aggregated across bootstrap samples and varying degrees of the energy
penalty \(\lambda_{\text{egy}}\) used for \emph{ECCo} at test time.
Standard errors for accuracy are bootstrapped from the test set.}

@@

\section{Discussion}\label{sec-discussion}

@@

\emph{Interventions on features have implications for fairness.} We
provide a method to modify the sensitivity of a model to certain
features, which can be misused by enforcing explanations based on
features that are more difficult to modify by a (group of) decision
subjects. Such abuse could result in an unfairly assigned burden of
recourse \citep{sharma2020certifai}, threatening the equality of
opportunity \citep{bell2024fairness}. Also, even if all immutable
features are protected, there may exist proxies that are theoretically
mutable, but preserve sufficient information about the principals to
hinder these protections. Indeed, deciding on the actionability of
features remains a major open challenge in the AR literature
\citep{venkatasubramanian2020philosophical}.

\emph{CT increases the training times.} Like adversarial training, CT is
more resource-intensive than conventional regimes. Higher numbers of CEs
improve the quality of the learned representations but they also
increase the number of computations. As our codebase is not performance
optimized, grids of around 300 settings for the largest datasets in our
experiments took up to four hours using 34 2GB CPUs (see supplementary
appendix). Other than optimization, three factors mitigate this effect:
(1) CT yields itself to parallel execution; (2) it amortizes the cost of
CEs for the training samples; and (3) it can be used to fine-tune
conventionally-trained models.

We also highlight three key directions for future research. Firstly, it
is an interesting challenge to extend CT beyond classification settings.
Our formulation relies on the distinction between non-target class(es)
and target class(es), requiring the output space to be discrete. Thus,
it does not apply to ML tasks where the change in outcome cannot be
readily discretized. Focus on classification is a common choice in
research on CEs and AR; other settings have attracted some interest,
e.g., regression \citep{spooner2021counterfactual}, but there is little
consensus how to robustly extend the notion of CEs.\\
Secondly, CT is susceptible to training instabilities, as also
recognized for the related JEMs \citep{grathwohl2020your}. CT is exposed
to two sources of instabilities: (1) the energy-based contrastive
divergence term, \(\text{div}(\cdot)\), in Equation~\ref{eq-div}, and
(2) the underlying explainers. We find several promising ways to
mitigate this problem: regularizing energy (\(\lambda_{\text{reg}}\)),
generating sufficiently many counterfactuals during each epoch, and
including only mature counterfactuals in \(\text{div}(\cdot)\).\\
Finally, we believe that it is possible to considerably improve
hyperparameter selection procedures, and thus performance. We have
relied exclusively on grid searches, but future work could benefit from
more sophisticated approaches.

\section{Conclusions}\label{sec-conclusions}

State-of-the-art machine learning models are prone to learning complex
representations that cannot be interpreted by humans. Existing
explainability solutions cannot guarantee that explanations agree with
these learned representation. As a step towards addressing this
challenge, we introduce counterfactual training, a novel training regime
that incentivizes highly-explainable models. Through extensive
experiments, we demonstrate that CT satisfies this objective while
promoting robustness and preserving the predictive performance of
models. Explanations generated from CT-based models are both more
plausible (compliant with the underlying data-generating process) and
more actionable (compliant with user-specified mutability constraints),
and thus meaningful to their recipients. We also show that CT can be
used to fine-tune conventionally-trained models and achieve similar
gains. Lastly, our work highlights the value of simultaneously improving
models and their explanations.

### 2025-07-17 paper/sections/conclusion.qmd
Trim to 7 pages
Draft adjustments in discussions/conclusions

To conclude, state-of-the-art machine learning models are prone to learning complex representations that cannot be interpreted by humans. Existing explainability solutions cannot guarantee that explanations agree with the model's learned representation of data. As a step towards addressing this challenge, we introduced counterfactual training, a novel training regime that incentivizes highly-explainable models. Our approach leads to explanations that are both plausible (compliant with the underlying data-generating process) and actionable (compliant with user-specified mutability constraints), and thus meaningful to their recipients. Through extensive experiments, we demonstrated that CT satisfies its objective while promoting robustness and preserving the predictive performance of models. It can also be used to fine-tune conventionally-trained models and achieve similar gains. Thus, our work showcases that it is practical to simultaneously improve models and their explanations.

>>>>

# Conclusions {#sec-conclusions}

State-of-the-art machine learning models are prone to learning complex representations that cannot be interpreted by humans. Existing explainability solutions cannot guarantee that explanations agree with these learned representation. As a step towards addressing this challenge, we introduce counterfactual training, a novel training regime that incentivizes highly-explainable models. Through extensive experiments, we demonstrate that CT satisfies this objective while promoting robustness and preserving the predictive performance of models. Explanations generated from CT-based models are both more plausible (compliant with the underlying data-generating process) and more actionable (compliant with user-specified mutability constraints), and thus meaningful to their recipients. We also show that CT can be used to fine-tune conventionally-trained models and achieve similar gains. Lastly, our work highlights the value of simultaneously improving models and their explanations.

### 2025-07-17 paper/sections/discussion.qmd
Draft adjustments in discussions/conclusions

# Conclusions {#sec-discussion}

@@

*Interventions on features have implications for fairness.* We provide a tool that allows practitioners to modify the sensitivity of a model with respect to certain features. Model owners can use our solution to support the fair and equitable treatment of decision subjects, but they could also misuse it by enforcing explanations based on features that are more difficult to modify by some (group of) individuals. When used irresponsibly, CT could result in an unfairly assigned burden of recourse [@sharma2020certifai], threatening the equality of opportunity [@bell2024fairness]. Additionally, CT requires mutability constraints for the features considered by the model. Even if all immutable features are protected, there may exist proxies that are mutable, and hence should not be protected, but preserve sufficient information about the principals to hinder these protections. Deciding on actionability is still a major open challenge in the AR literature [@venkatasubramanian2020philosophical] impacting the capacity of CT to fulfill its intended goal.

*CT increases the training times.* Like adversarial training, CT is more resource-intensive than conventional regimes. Higher numbers of CEs improve the quality of learned representations but they also increase the number of computations. As our codebase is not performance optimized, grids of 270 settings for the largest datasets in our experiments took up to four hours using 34 2GB CPUs (see supplementary appendix). Other than optimization, three factors mitigate this effect: (1) CT yields itself to parallel execution; (2) it amortizes the cost of CEs for the training samples; and (3) it can be used to fine-tune conventionally-trained models.

We also highlight three important directions for future research. Firstly, it is an interesting challenge to extend CT beyond classification settings. Our formulation relies on the distinction between non-target class(es) $y^{-}$ and target class(es) $y^{+}$ to generate counterfactuals through @eq-obj. While $y^{-}$ and $y^{+}$ can be arbitrarily defined, CT requires the output space $\mathcal{Y}$ to be discrete. Thus, it does not apply to ML tasks where the change in outcome cannot be readily discretized. Focus on classification is a common choice in research on CEs and AR. Other settings have attracted some interest, e.g., regression [@spooner2021counterfactual], but there is little consensus how to robustly extend the notion of CEs.  
Secondly, CT is susceptible to training instabilities. This problem has been recognized for JEMs [@grathwohl2020your] and even though we depart from the SGLD sampling, we still encounter variability in outcomes. CT is exposed to two potential sources of instabilities: (1) the energy-based contrastive divergence term, $\text{div}(\cdot)$, in @eq-div, and (2) the underlying explainers. We find several promising ways to mitigate this problem: regularizing energy ($\lambda_{\text{reg}}$), generating sufficiently many counterfactuals during each epoch, and including only mature counterfactuals in $\text{div}(\cdot)$.  
Finally, we believe that it is possible to considerably improve hyperparameter selection procedures, and thus performance. We have relied exclusively on grid searches, but future research could benefit from more sophisticated approaches.

>>>>

# Discussion {#sec-discussion}

@@

*Interventions on features have implications for fairness.* We provide a method to modify the sensitivity of a model to certain features, which can be misused by enforcing explanations based on features that are more difficult to modify by a (group of) decision subjects. Such abuse could result in an unfairly assigned burden of recourse [@sharma2020certifai], threatening the equality of opportunity [@bell2024fairness]. Also, even if all immutable features are protected, there may exist proxies that are theoretically mutable, but preserve sufficient information about the principals to hinder these protections. Indeed, deciding on the actionability of features remains a major open challenge in the AR literature [@venkatasubramanian2020philosophical].

*CT increases the training times.* Like adversarial training, CT is more resource-intensive than conventional regimes. Higher numbers of CEs improve the quality of the learned representations but they also increase the number of computations. As our codebase is not performance optimized, grids of around 300 settings for the largest datasets in our experiments took up to four hours using 34 2GB CPUs (see supplementary appendix). Other than optimization, three factors mitigate this effect: (1) CT yields itself to parallel execution; (2) it amortizes the cost of CEs for the training samples; and (3) it can be used to fine-tune conventionally-trained models.

We also highlight three key directions for future research. Firstly, it is an interesting challenge to extend CT beyond classification settings. Our formulation relies on the distinction between non-target class(es) and target class(es), requiring the output space to be discrete. Thus, it does not apply to ML tasks where the change in outcome cannot be readily discretized. Focus on classification is a common choice in research on CEs and AR; other settings have attracted some interest, e.g., regression [@spooner2021counterfactual], but there is little consensus how to robustly extend the notion of CEs.  
Secondly, CT is susceptible to training instabilities, as also recognized for the related JEMs [@grathwohl2020your]. CT is exposed to two sources of instabilities: (1) the energy-based contrastive divergence term, $\text{div}(\cdot)$, in @eq-div, and (2) the underlying explainers. We find several promising ways to mitigate this problem: regularizing energy ($\lambda_{\text{reg}}$), generating sufficiently many counterfactuals during each epoch, and including only mature counterfactuals in $\text{div}(\cdot)$.  
Finally, we believe that it is possible to considerably improve hyperparameter selection procedures, and thus performance. We have relied exclusively on grid searches, but future work could benefit from more sophisticated approaches.

### 2025-07-17 paper/sections/intro.qmd
Draft adjustments in discussions/conclusions

Specifically, **we propose Counterfactual Training (CT)**: a novel training regime explicitly geared towards improving the explanatory capacity of models that, in high-level terms, we define as the extent to which valid counterfactual explanations derived for a model can be deemed plausible with respect to the underlying data and global actionability constraints (we refine this notion in Def. \ref{def-explainability}). To the best of our knowledge, Counterfactual Training represents the first attempt to achieve more explainable models by employing counterfactual explanations already in the training phase.

The remainder of this manuscript is structured as follows. @sec-lit presents related work, focusing on the link between AEs and CEs. Then follow our two principal contributions. In @sec-method, we introduce our methodological framework and show theoretically that it can be employed to enforce global actionability constraints. In @sec-experiments, through extensive experiments, we empirically demonstrate that CT substantially improves explainability and positively contributes to the robustness of trained models without sacrificing predictive performance. Finally, in @sec-discussion, we discuss open challenges and conclude that CT is a promising approach towards making opaque models more trustworthy.

>>>>

Specifically, **we propose Counterfactual Training (CT)**: a novel training regime explicitly geared towards improving the explanatory capacity of models that, in high-level terms, we define as the extent to which valid explanations derived for a model can be deemed plausible with respect to the underlying data and global actionability constraints (we refine this notion in Def. \ref{def-explainability}). For simplicity, we refer to models with high explanatory capacity as *explainable*. To the best of our knowledge, Counterfactual Training represents the first attempt to achieve more explainable models by employing counterfactual explanations already in the training phase.

The remainder of this manuscript is structured as follows. @sec-lit presents related work, focusing on the link between AEs and CEs. Then follow our two principal contributions. In @sec-method, we introduce our methodological framework and show theoretically that it can be employed to enforce global actionability constraints. In @sec-experiments, through extensive experiments, we empirically demonstrate that CT substantially improves explainability and positively contributes to the robustness of trained models without sacrificing predictive performance. In @sec-discussion, we discuss open challenges and, in @sec-conclusion, conclude that CT is a promising approach towards making opaque models more trustworthy.
